---
title: "Homework 1"
output: html_notebook
---

**Rules**:

1. **The submission deadline is 26.10.2020, 23:59:59**. Your homework should be submitted to `hwdsehse@gmail.com`. **No late submissions will be accepted**;

2. Every group should submit one `.Rmd` file (you should use this file to write up your homework). Submitted file should be named: `hw_1_vjacisens_ntoropov.Rmd` for the file submitted by Vitalijs Jascisens and Nikita Toropov;

3. **No plagiarism**: all sources (both offline and online have to be acknowledged). Cheating on any part of the assignment results in a **grade of zero** for the entire assignment;

4. We will not be checking your code. All problems have precise answers. Only the correctness of the answer will be graded.

**Name and Surname**:

1. Maria Adshead (3);

# Problem 1: XML Parsing (20 points)

**Data**: In this problem you will work with the data stored in the folder `data_p_1`.

**Rules:**
Functions that you are allowed to use in this problem:

  * Functions from base `R`;
  
  * Functions from `xml2` library;
  
  * Functions from `stringr` library;
  
  * Functions from `data.table` library.

**Question**: Write a function that for a given `.xml` file does the following:

1. Calculates the sum of `СумДоход` across all firms within a file;

2. Calculates the sum of `СумРасход` across all firms within a file;

3. The function should return a vector of length 2. The first element should be `СумДоход` across all firms within a file, the second element should be `СумРасход` across all firms within a file. When calculting both sums set `na.rm=TRUE`.

```{r}
#Load needed libraries to use in this project:
library(xml2)
library(stringr)
library(data.table)

#Define the working directory from which you are working:
dir_p_1<-"/Users/vitalijs/Downloads/data_p_1"

#Set the working directory:
setwd(dir_p_1)

#Your solution comes here:
function_extract_info_fns<-function(x){
    x<-read_xml(x)
  dt<-data.table("СумДоход"=numeric(),"СумРасход"=numeric())
  alldocs<-xml_children(x)
  for (i in 2:length(alldocs)){
    d<-xml_attrs(xml_child(xml_child(x, i), 2))[["СумДоход"]]
    r<-xml_attrs(xml_child(xml_child(x, i), 2))[["СумРасход"]]
    dt<-rbind(dt, data.table("СумДоход"= as.numeric(d),"СумРасход" = as.numeric(r)))
  }
  sumdohod<-sum(dt[,c("СумДоход")],na.rm=TRUE)
  sumrashod<-sum(dt[,c("СумРасход")],na.rm=TRUE)
  vec<-c(sumdohod,sumrashod)
  return(vec)
}
#Sources:
#https://www.rdocumentation.org/packages/xml2/versions/1.3.2
#https://www.rdocumentation.org/packages/xml2/versions/1.3.2/topics/xml_children
#https://stackoverflow.com/questions/29711377/how-do-i-initialize-an-empty-or-example-data-table
```

# Problem 2: More XML Parsing (40 points)

**Data**: In this problem you will work with the data stored in folders `data_p_2_1` and `data_p_2_2`.

**Rules:**
Functions that you are allowed to use in this problem:

  * Functions from base `R`;
  
  * Functions from `xml2` library;
  
  * Functions from `stringr` library;
  
  * Functions from `data.table` library.


**Questions**:

1. We start with the data `data_p_2_1`.Write a function that for a given `.xml` file does the following:
  
  * For each firm extracts:
  
    * `ИННФЛ`;
    * `ИННЮЛ`;
    * `КодРегион`;
    * `СвОКВЭДОсн КодОКВЭД`: here you should be careful, I want you take the value with `СвОКВЭДОсн` and not with `СвОКВЭДДоп`;
    * If there are multiple fields within a tag, combines these fields using `&&&&`;
    * If any of above fields are missing the resulting value for that field should be `NA`.
  
  * Combines (by row) this information into the data.table;
  
  * Writes this data.table into a `.csv.gz` file (please use option `compress = "gzip"`). The name of the file should be the same as the name of an `.xml` file except that an `.xml` extension should be replaced by a `.csv.gz` extension. You should write the data directly to `data_p_2_1` folder.

```{r}

#Load needed libraries to use in this project:
library(xml2)
library(stringr)
library(data.table)

#Define the working directory from which you are working:
dir_p_2_1<-"/Users/vitalijs/Desktop/data_p_2_1"

#Set the working directory:
setwd(dir_p_2_1)

#Your solution comes here:
function_process_one_xml<-function(x){
  newname<-str_replace(x,'.xml','.csv.gz')
  newnamegz<-str_glue('~/Downloads/data_p_2_1/',newname)
  x<-read_xml(x)
  alldoc<-xml_find_all(x,'//Документ')
  data<-data.table(
    "ИННЮЛ" = character(),
    "ИННФЛ" = character(),
    "КодРегион" = character(),
    "КодОКВЭД" = character()
  )
  for (i in (1:length(alldoc))){
    childrenlist<-xml_name(xml_children(alldoc[i]))
    if (is.element('ОргВклМСП',childrenlist)&(is.element('ИПВклМСП',childrenlist))){
      innul<-xml_attrs(xml_child(alldoc[[i]], 'ОргВклМСП'))[["ИННЮЛ"]]
      innfl<-xml_attrs(xml_child(alldoc[[i]], 'ИПВклМСП'))[["ИННФЛ"]]
    } else if (is.element('ОргВклМСП',childrenlist)){
      innul<-xml_attrs(xml_child(alldoc[[i]], 'ОргВклМСП'))[["ИННЮЛ"]]
      innfl<-NA
    } else if (is.element('ИПВклМСП',childrenlist)){
      innfl<-xml_attrs(xml_child(alldoc[[i]], 'ИПВклМСП'))[["ИННФЛ"]]
      innul<-NA
    } else {
      innfl<-NA
      innul<-NA
    }
    if (is.element('СведМН',childrenlist)) {
      kod<-xml_attrs(xml_child(alldoc[[i]], 'СведМН'))[["КодРегион"]]
    } else {
      kod<-NA
    }
    if (is.element('СвОКВЭД',childrenlist)) {
      childrenlist2<-xml_name(xml_children(xml_child(alldoc[[i]],'СвОКВЭД')))
      if (is.element('СвОКВЭДОсн',childrenlist2)) {
        kodokved<-xml_attrs(xml_child(xml_child(alldoc[[i]], 'СвОКВЭД'), 'СвОКВЭДОсн'))[["КодОКВЭД"]]
      }
    } else {
      kodokved<-NA
    }
    data<-rbind(data, data.table("ИННЮЛ" = innul,"ИННФЛ" = innfl ,"КодРегион" = kod,"КодОКВЭД" = kodokved))
  }
  fwrite(data, file=newnamegz,compress = 'gzip')
}
#Sources:
#tutorials
#https://stackoverflow.com/questions/29711377/how-do-i-initialize-an-empty-or-example-data-table
#https://www.c-sharpcorner.com/article/r-data-frame-operations-adding-rows-removing-rows-and-merging-two-data-frame/
#https://www.rdocumentation.org/packages/xml2/versions/1.3.2/topics/xml_name
#https://stackoverflow.com/questions/1169248/test-if-a-vector-contains-a-given-element
```


2. Next, we turn our attention to the data `data_p_2_2`. Write a function that does the following:

  * Reads an `.rds` file;
  
  * Extracts from this file following fields:
    
    * `ДатаВклМСП`;
    
    * `КодРегион`;
    
    * `ИННЮЛ`;
    
    * `ИННФЛ`.

```{r}
#Load needed libraries to use in this project:
library(xml2)
library(stringr)
library(data.table)

#Define the working directory from which you are working:
dir_p_2_2<-"/Users/vitalijs/Desktop/data_p_2_2"

#Set the working directory:
setwd(dir_p_2_2)

#Your solution comes here:
function_process_one_rds<-function(x){
  x<-readRDS(x)
  dt<-data.table(x)
  if (is.element('ИННФЛ',colnames(dt))&is.element('ИННЮЛ',colnames(dt))){
    dt<-dt[,c('ДатаВклМСП','КодРегион','ИННЮЛ','ИННФЛ')]
  } else if (is.element('ИННФЛ',colnames(dt))){
    dt<-dt[,c('ДатаВклМСП','КодРегион','ИННФЛ')]
  } else if (is.element('ИННЮЛ',colnames(dt))){
    dt<-dt[,c('ДатаВклМСП','КодРегион','ИННЮЛ')]
  }
}
```

3. Use `function_process_one_rds` to combine information from all files into the `data.table` and call it `data`;

4. Using `data` answer following questions:

  * Which region has:
    
    * The largest number of `ИННЮЛ`;
    
    * The largest number of `ИННФЛ`.
  
  * On which date were the maximum number of firms created.
  
```{r}
data<-data.table(
  "ДатаВклМСП" = character(),
  "КодРегион" = character(),
  "ИННЮЛ" = character(),
  "ИННФЛ" = character()
)

for (i in 1:length(filenames)){
  d<-function_process_one_rds(filenames[i])
  data<-rbind(data,d, fill=TRUE)
}
data<-data[,c(5,6,7,8)]

#region for largest number of innul

innul_dt<-data[,c('КодРегион','ИННЮЛ')]
innul_dt<-na.omit(innul_dt)

unique_reg_innul<-list(unique(innul_dt[,as.character(c('КодРегион'))]))
grouped_reg_innul<-groupingsets(innul_dt,j = list(count=.N),by='КодРегион',sets = unique_reg_innul)
max_innul_count<-max(grouped_reg_innul$count)

#region for largest number of innfl

innfl_dt<-data[,c('КодРегион','ИННФЛ')]
innfl_dt<-na.omit(innfl_dt)

unique_reg_innfl<-list(unique(innfl_dt[,as.character(c('КодРегион'))]))
grouped_reg_innfl<-groupingsets(innfl_dt,j = list(count=.N),by='КодРегион',sets = unique_reg_innfl)
max_innfl_count<-max(grouped_reg_innfl$count)

#on which date were the maximum number of firms created

firm_creation<-data[,c('ДатаВклМСП','ИННЮЛ')]
firm_creation<-na.omit(firm_creation)

unique_date<-list(unique(firm_creation[,as.character(c('ДатаВклМСП'))]))
grouped_date<-groupingsets(firm_creation,j = list(count=.N),by='ДатаВклМСП',sets = unique_date)
max_firms<-max(grouped_date$count)

region_max_nr_firms<-grouped_reg_innul$КодРегион[grouped_reg_innul$count==max_innul_count]
region_max_nr_ind<-grouped_reg_innfl$КодРегион[grouped_reg_innfl$count==max_innfl_count]
date_max_firms<-grouped_date$ДатаВклМСП[grouped_date$count==max_firms]
#region_max_nr_firms<-'77'
#region_max_nr_ind<-'77'
#date_max_firms<-'01.08.2016'

#Sources: (+some of the ones provided above)
#https://rdrr.io/cran/data.table/man/groupingsets.html
#https://stackoverflow.com/questions/7562284/extracting-unique-rows-from-a-data-table-in-r
#https://stackoverflow.com/questions/19869145/count-number-of-records-and-generate-row-number-within-each-group-in-a-data-tabl
```

# Problem 3: Playing With JSON (20 points)

**Data**: In this problem you will work with the data `data_p_3.json`.

**Rules:**
Functions that you are allowed to use in this problem:

  * Functions from base `R`;
  
  * Functions from `xml2` library;
  
  * Functions from `stringr` library;
  
  * Functions from `data.table` library;
  
  * Functions from `jsonlite` library;
  
  * Functions from `rjson` library.
  
**Questions**:

0. Read the data.

```{r}
#Clear everything and load needed libraries:
rm(list=ls())
library(data.table)
library(jsonlite)

#Define directory where the data is stored:
dir_data<-"/users/vitalijs/Desktop"

#Read the data:
setwd(dir_data)
data<-lapply(readLines("data_p_3.json"),fromJSON)
```

1. As you can see `data` is a complicated list. Each element of the list contains information on one ngo. Using `data` answer following questions:

  * Which region has the most `ngos`;
  
  * Which `ngo` has received most governmental contracts (as measured by `contracts44$totalSum`). Provide the `fullName` of this ngo.

```{r}
dt<-data.table('ngo'=character(),'regionName'=character())

for (i in 1:length(data)){
  name<-data[[i]]$regionName
  ngo<-i
  dt<-rbind(dt, data.table("ngo" = i,"regionName" = name))
}

dt<-na.omit(dt)

unique_region<-list(unique(dt[,as.character(c('regionName'))]))
grouped_region<-groupingsets(dt,j = list(count=.N),by='regionName',sets = unique_region)
max_ngos<-max(grouped_region$count)

dt<-data.table('ngo'=character(),'contracts'=numeric(),'fullName'=character())

for (i in 1:length(data)){
  contr<-data[[i]]$incomeDetail$contracts44$totalSum
  name<-data[[i]]$fullName
  ngo<-i
  dt<-rbind(dt, data.table("ngo" = i,"contracts" = as.numeric(contr),'fullName'=name))
}

max_contract<-max(dt$contracts)

region_max_ngos<-grouped_region$regionName[grouped_region$count==max_ngos]
ngo_max_contract<-dt$fullName[dt$contracts==max_contract]

#region_max_ngos<-"Москва"
#ngo_max_contract<-"МУНИЦИПАЛЬНОЕ ПРЕДПРИЯТИЕ ГОРОДА ВЛАДИВОСТОКА \"ВЛАДИВОСТОКСКОЕ ПРОИЗВОДСТВЕННОЕ ОБЪЕДИНЕНИЕ ПАССАЖИРСКОГО АВТОТРАНСПОРТА №1\""
region_max_ngos
ngo_max_contract
```
  
2. Read the file `data_p_3.json` using `stream_in()` function from the `jsonlite` library. Using this data calculate:

  * Which region has the most `ngos`;
  
  * Which `ngo` has received most governmental contracts (as measured by `contracts44$totalSum`). Provide the `fullname` of this ngo.

```{r}
#Sources:
#https://www.rdocumentation.org/packages/jsonlite/versions/1.7.1/topics/stream_in%2C%20stream_out
data_stream<-stream_in(file("data_p_3.json"))

income<-data_stream[,'incomeDetail']
contracts44<-income[,'contracts44']
sum<-contracts44[,'totalSum']
dt<-data.table('fullName'=data_stream[,'fullName'],'regionName'=data_stream[,'regionName'],totalSum=as.numeric(sum))

unique_region<-list(unique(dt[,as.character(c('regionName'))]))
grouped_region<-groupingsets(dt,j = list(count=.N),by='regionName',sets = unique_region)
max_ngos<-max(grouped_region$count)

max_contract<-max(dt$totalSum)

region_max_ngos<-grouped_region$regionName[grouped_region$count==max_ngos]
ngo_max_contract<-dt$fullName[dt$totalSum==max_contract]

#region_max_ngos<-"Москва"
#ngo_max_contract<-"МУНИЦИПАЛЬНОЕ ПРЕДПРИЯТИЕ ГОРОДА ВЛАДИВОСТОКА \"ВЛАДИВОСТОКСКОЕ ПРОИЗВОДСТВЕННОЕ ОБЪЕДИНЕНИЕ ПАССАЖИРСКОГО АВТОТРАНСПОРТА №1\""
region_max_ngos
ngo_max_contract
```


# Problem 4: Regular Expressions (20 points)

0. Read the data.
```{r}
#Load needed libraries to use in this project:
library(xml2)
library(stringr)
library(data.table)

#Define the working directory from which you are working:
dir_p_4<-"/users/vitalijs/Desktop"

#Load the data:
data<-read_html("plc_list.txt")
data<-xml_text(data)

#The data is displayed nicely here:
#https://home.treasury.gov/system/files/126/plc_list.txt
```
**Rules:**
Functions that you are allowed to use in this problem:

  * Functions from base `R`;
  
  * Functions from `xml2` library;
  
  * Functions from `stringr` library;
  
  * Functions from `data.table` library.
  
**Questions:**

1. Count the total number of individuals that are included in the list. The list starts with `ABDEL JAWAD, Nasir Abdullah Odeh` and ends with `ZEIDAN, Abdul Rahman Fahmi`. To solve this problem you should assume the following:

* You should create paragraphs using `data`;
* Each paragraph is an individual;
* The name of an individual cannot start with `Directive`. I.e., you should remove all paragraphs that start with `Directive`;
* The name of an individual cannot start with `_`. I.e., you should remove all paragraphs that start with `_`;
* The name of an individual cannot start with `For further information`. I.e., you should remove all paragraphs that start with `For further information`;
* The name of an individual cannot start with `9/16/2016`. I.e., you should remove all paragraphs that start with `9/16/2016`.

```{r}
s<-str_split(data,'\\n\n')
d<-s[[1]]
newlist<-list()

for (i in 1:length(d)){
  if (str_detect(d[i],"^Directive")==F&str_detect(d[i],"^_")==F&str_detect(d[i],"^For further information")==F&str_detect(d[i],"^9/16/2016")==F&str_detect(d[i],"^This publication")==F){
    newlist<-c(newlist, d[i])
  }
}
# Answer comes here:
nr_total<-length(newlist)
nr_total
#Final answer:  
#nr_total=161
```

2. Write a function that extracts the name of an individual given one paragraph. To solve this problem you should proceed as follows:

* Your function should extract everything (including `\n`) from the beginning of the string until first `(a.k.a`;

* If the paragraph does not contain `(a.k.a` then your function should do the following:

  1. First extract everything (including `\n`) from the beginning of the string until first `;` (or the the whole string if there are no `;` in the string);
  
  2. Extract everything until the second `,`. If there are less than 2 comas then your function should return string from the previous step.
  
**Examples**:
  
  * `ZAHHAR, Mahmoud Khaled (a.k.a. AL ZAHAR, Mahmoud Khaled); DOB 1945\n(individual) [NS-PLC]`: your function should return `ZAHHAR, Mahmoud Khaled`;
  
  * `"SALEH, Mariam Mahmoud, Ramallah, West Bank, Palestinian; DOB 1952\n(individual) [NS-PLC]."`: your function should return `SALEH, Mariam Mahmoud`;
  
  * `SA'DAT, Ahmad; PFLP (individual) [NS-PLC]`: your function should return `SA'DAT, Ahmad`;

* Finally, your function should remove (please proceed in this order):

  1. `(a.k.a` at the end of the string;
  2. `,` at the end of the string;
  3. `;` at the end of the string;
  4. All whitespaces at the end and beginning of the string.


```{r}
#Your code comes here:
function_extract_name<-function(x){
    if (str_detect(x,'a.k.a')==TRUE){
    a<-str_locate(x, fixed('('))
    name<-substr(x,1,(a-2))
  } else if (str_detect(x,';')==TRUE){
    a<-str_locate(x,';')
    name0<-substr(x,1,(a-1))
    commas<-str_locate_all(name0, fixed(','))[[1]]
    if (length(commas)<=2){
      name<-name0
    } else if (length(commas)>2){
      name<-substr(name0,1,(commas[3]-1))
    }
  } else if (str_detect(x,';')==FALSE){
    name0<-x
    commas<-str_locate_all(name0, fixed(','))[[1]]
    if (length(commas)<=2){
      name<-name0
    } else if (length(commas)>2){
      name<-substr(name0,1,(commas[3]-1))
    }
  }
  return(name)
}
#Sources:
#http://dept.stat.lsa.umich.edu/~jerrick/courses/stat701/notes/stringmanip.html
```

3. Write a function that extracts the year of birth of a given individual (given the paragraph). If the year of birth is not present your function should return `NA`.

```{r}
#Your code comes here:
function_extract_yob<-function(x){
    if (str_detect(x,fixed('9'))==TRUE){
    a<-str_locate(x,fixed('9'))
    yob<-substr(x,(a-1),(a+2))
  } else {
    yob<-NA
  }
  return(yob)
}
```
